---
title: "Modular design and software development best practices in R"
institute: "Odum Institute<br/>University of North Carolina at Chapel Hill"
format:
    revealjs:
        theme: [default, unc.scss]
        width: 1920
        height: 1080
        logo: unc_logo.svg
        slide-number: true
        footer: "https://go.unc.edu/modulardesign"
execute:
  include: true
  echo: true
---

```{r}
#| include: false
#| echo: false

library(tidyverse)

data = read_csv("../data/air_sample.csv")

market_ids = read_csv("../data/L_CITY_MARKET_ID.csv")
data = left_join(data, rename(market_ids, OriginCity="Description"), by=c(OriginCityMarketID="Code"))
data = left_join(data, rename(market_ids, DestCity="Description"), by=c(DestCityMarketID="Code"))

carriers = read_csv("../data/L_CARRIERS.csv")
data = left_join(data, rename(carriers, OperatingCarrierName="Description"), by=c(OpCarrier="Code"))
data = left_join(data, rename(carriers, TicketingCarrierName="Description"), by=c(TkCarrier="Code"))
```

## About me

- Assistant Professor of City and Regional Planning
    - Research focus: transportation modeling and simulation
- Quantitative Methods Consultant at Odum
- Three years as a software developer before graduate school

## Why this matters

- Code is increasingly common in research
- Most academics receive little to no formal training in software development
- This can lead to
    - Inefficient development (takes too much time)
    - Errors in code
    - Non-reproducible analyses

## What we'll be doing today

- Hands-on exercise converting a set of R scripts a modular and tested format
- Two scripts that analyze data on airline tickets from the Bureau of Transportation Statistics
- Script repository (clone this now): [https://github.com/mattwigway/OdumModularDesignR](https://github.com/mattwigway/OdumModularDesignR)
- These slides: [https://projects.indicatrix.org/OdumModularDesignR](https://projects.indicatrix.org/OdumModularDesignR)

## The two scripts

- Both load and clean the air ticket data
- One analyzes the busiest air routes in the country
- One looks at airline "fortress hubs"---which airports have the most concentrated service from a single airline

## What we'll address

- Duplicated code
- Literal numeric values
- Automated tests
- Package version incompatibilities

## What we won't address

- Version control
    - Odum offers separate Git and Github class

## `air_analysis.R`

- Hands-on walkthrough of file

## `airline_analysis.R`

- Hands-on walkthrough of file

## Duplicated code (don't repeat yourself)

- Common software development rule: don't repeat yourself (DRY)
- In `air_analysis.R` we do the same analysis twice, once by airport and once by city
- We can eliminate this duplicated code using a _function_

## Functions

- A function is a reusable piece of code that can be executed many times
- Functions separate the _definition_ of code from the _execution_
- You use functions all the time
    - For instance, `sum()` is a function

## Anatomy of a function

- Functions are R code that is given a name to be referred to later
- Functions have three main components
    - Body
        - The main R code of the function
    - Arguments
        - Values that can be specified each time the function is used
    - Return value
        - The result of the function

## Anatomy of a function


```{r}
calculate_cylinder_volume = function (radius, height) {
    base_area = pi * radius ^ 2
    volume = base_area * height
    return(volume)
}
```

::: {.incremental}
- _radius_ and _height_ are arguments
- Next two lines are body (the actual calculations)
- _volume_ is the return value
    - `return` ends the function, any code after `return` will not be run
:::

## Using functions

- Code above _defines_ the function
- It doesn't run any of the function code
- Function code is run when the function is _called_

. . . 

```{r}
calculate_cylinder_volume(1, 2)
```

## Defining functions in `air_analysis.R`

- The analysis is repeated twice, once by airport pairs and once by city pairs
- We will re-write this analysis into a function, and use the same code twice

## Defining functions in `air_analysis.R`

```{r}
#| output: false
pairs = group_by(data, Origin, Dest) |> summarize(Passengers=sum(Passengers),
    distance_km=first(Distance) * 1.609)
arrange(pairs, -Passengers)

pairs = mutate(pairs,
    airport1 = if_else(Origin < Dest, Origin, Dest),
    airport2 = if_else(Origin < Dest, Dest, Origin)
    )
pairs = group_by(pairs, airport1, airport2) |> summarize(
    Passengers=sum(Passengers),
    distance_km=first(distance_km)
)
arrange(pairs, -Passengers)
```

## Step one: make it a function

- Wrap code in `function () {...}` and add a return value
    - If you don't have a `return()` R will return the result of the last line

```{r}
#| output: false
busiest_routes = function () {
  # Now, we can see what the most popular air route is, by summing up the number of
  # passengers carried.
  pairs = data |>
    group_by(Origin, Dest) |>
    summarize(Passengers = sum(Passengers), distance_km = first(Distance) * 1.609) |>
    ungroup()
  
  # we see that LAX-JFK (Los Angeles to New York Kennedy) is represented separately
  # from JFK-LAX. We'd like to combine these two. Create airport1 and airport2 fields
  # with the first and second airport in alphabetical order.
  pairs = pairs |>
    mutate(
      airport1 = if_else(Origin < Dest, Origin, Dest),
      airport2 = if_else(Origin < Dest, Dest, Origin)
    ) |>
    group_by(airport1, airport2) |>
    summarize(Passengers = sum(Passengers), distance_km = first(distance_km))
  
  arrange(pairs, -Passengers)
}
```

## Running our function

- Running the code on the previous slide defines the function, but does not run it
- We have to _call_ the function to run it

```{r}
busiest_routes()
```

## Adding arguments

- We want to make the `Origin` and `Dest` column names arguments to the function
- It's also standard to make the data itself and anything else you refer to inside the function arguments

## Adding arguments

```{r}
#| output: false
busiest_routes = function (dataframe, origcol, destcol) {
  # Now, we can see what the most popular air route is, by summing up the number of
  # passengers carried.
  pairs = dataframe |>
    group_by(origcol, destcol) |>
    summarize(Passengers = sum(Passengers), distance_km = first(Distance) * 1.609) |>
    ungroup()
  
  # we see that LAX-JFK (Los Angeles to New York Kennedy) is represented separately
  # from JFK-LAX. We'd like to combine these two. Create airport1 and airport2 fields
  # with the first and second airport in alphabetical order.
  pairs = pairs |>
    mutate(
      airport1 = if_else(origcol < destcol, origcol, destcol),
      airport2 = if_else(origcol < destcol, destcol, origcol)
    ) |>
    group_by(airport1, airport2) |>
    summarize(Passengers = sum(Passengers), distance_km = first(distance_km))
  
  arrange(pairs, -Passengers)
}
```

## Adding arguments

```{r}
#| error: true
busiest_routes(data, Origin, Dest)
```

## What happened?

- R tried to group by the columns `origcol` and `destcol`
- These columns don't exist
- We need to tell R that `origcol` and `destcol` aren't actually column names, but variables that contain column names
- Any place we refer to a variable containing a column name inside a function, use the "embracing operator" `{{{ variable }}}` to tell R to use the value contained by `variable` as the column name
    - This is only true when the variable name would not usually be in quotes (`" or '`)

## Adding arguments

```{r}
#| output: false
busiest_routes = function (dataframe, origcol, destcol) {
  # Now, we can see what the most popular air route is, by summing up the number of
  # passengers carried.
  pairs = dataframe |>
    group_by({{ origcol }}, {{ destcol }}) |>
    summarize(Passengers = sum(Passengers), distance_km = first(Distance) * 1.609) |>
    ungroup()
  
  # we see that LAX-JFK (Los Angeles to New York Kennedy) is represented separately
  # from JFK-LAX. We'd like to combine these two. Create airport1 and airport2 fields
  # with the first and second airport in alphabetical order.
  pairs = pairs |>
    mutate(
      airport1 = if_else({{ origcol }} < {{ destcol }}, {{ origcol }}, {{ destcol }}),
      airport2 = if_else({{ origcol }} < {{ destcol }}, {{ destcol }}, {{ origcol }})
    ) |>
    group_by(airport1, airport2) |>
    summarize(Passengers = sum(Passengers), distance_km = first(distance_km)) |>
    ungroup()
  
  arrange(pairs, -Passengers)
}
```

## Adding arguments

```{r}
busiest_routes(data, Origin, Dest)
```

## Using our function for cities

```{r}
busiest_routes(data, OriginCity, DestCity)
```

## Exercise: do the same in `airline_analysis.R`

- Write a function to compute market shares
- Use that function to compute market shares by operating carrier and ticketing carrier
- Bonus: also compute market shares for individual airports instead of cities

## Result

```{r}
# Now, we can compute the market shares
market_shares = function (dataframe, carrier, market) {
  mkt_shares = dataframe |>
    group_by({{ carrier }}, {{ market }}) |>
    summarize(Passengers = sum(Passengers)) |>
    group_by({{ market }}) |>
    mutate(market_share = Passengers / sum(Passengers), total_passengers = sum(Passengers)) |>
    ungroup()
  
  filter(mkt_shares, total_passengers > 1000) |> arrange(-market_share)
}

```

## Result

```{r}
market_shares(data, OperatingCarrierName, OriginCity)
```

## Result

```{r}
market_shares(data, TicketingCarrierName, OriginCity)
```

## Result

```{r}
market_shares(data, TicketingCarrierName, Origin)
```


## Symbolic constants

- We still have the value `1.609` in our function to convert miles to kilometers
- It's a bad practice to have literal values like these in your code
    - In a large project, value will end up spread over many files
    - Difficult to change (maybe you want to increase accuracy later by changing it to `1.609344`)
    - We can define a _constant variable_ in our file that contains the value `1.609`, and refer to it when needed
    - By convention, constants are all uppercase, and usually appear near the start of the file
    - R does not enforce variables being constant, you could change them in other code, but you shouldn't
    - Numbers used in many places can be prone to typos
    - R code will run fine and give the wrong answer if you accidentally entered `1.906`
    - But it will error if you entered `KIOLMETERS_TO_MILES`
    - Only one place where you have 

## Symbolic constants

- Near the top of your code file, add

```{r}
KILOMETERS_PER_MILE = 1.609
```

- Replace `1.609` in your function with `KILOMETERS_PER_MILE`

## Runtime checks: `stopifnot`

- A first step towards ensuring correctness is to add checks throughout your code
- In R, this is most easily done with the `stopifnot` function - this will error if whatever condition it checks is not true

## Runtime checks: `stopifnot`

- Let's make sure all of the passenger numbers are positive and non-NA
- Add these lines to the start of your `busiest_routes` function:
```{r}
#| eval: false
stopifnot(all(dataframe$Passengers >= 1))
stopifnot(all(!is.na(dataframe$Passengers)))
```

```{r}
#| include: false
busiest_routes = function (dataframe, origcol, destcol) {
  stopifnot(all(dataframe$Passengers >= 1))
  stopifnot(all(!is.na(dataframe$Passengers)))
  
  # Now, we can see what the most popular air route is, by summing up the number of
  # passengers carried.
  pairs = dataframe |>
    group_by({{ origcol }}, {{ destcol }}) |>
    summarize(Passengers = sum(Passengers), distance_km = first(Distance) * KILOMETERS_TO_MILES)
  arrange(pairs, -Passengers)
  
  # we see that LAX-JFK (Los Angeles to New York Kennedy) is represented separately
  # from JFK-LAX. We'd like to combine these two. Create airport1 and airport2 fields
  # with the first and second airport in alphabetical order.
  pairs = pairs |>
    mutate(
      airport1 = if_else({{ origcol }} < {{ destcol }}, {{ origcol }}, {{ destcol }}),
      airport2 = if_else({{ origcol }} < {{ destcol }}, {{ destcol }}, {{ origcol }})
    ) |>
    group_by(airport1, airport2) |>
    summarize(Passengers = sum(Passengers), distance_km = first(distance_km)) |>
    ungroup()
  
  arrange(pairs, -Passengers)
}
```

## Confirming that errors occur

```{r}
#| error: true
busiest_routes(mutate(data, Passengers=-1), Origin, Dest)
```

## All data manipulation should be in code

- I _never_ modify original data files, and rarely save new ones
- I always write code or functions to make all the changes/cleaning I need
- I run that function to load the data each time
- This way I can always see what has been done to data

## Avoiding absolute file paths

- Often, R code will include something like
    ```r
    read_csv("C:/Users/mwbc/Documents/Analysis/NHTS_2022/trips.csv")
    ```
- or
    ```r
    setwd("C:/Users/mwbc/Documents/Analysis/NHTS_2022")
    ```
- These are both fragile; if you share the code with someone else they won't work
- We want to avoid absolute file paths at all times

## Avoiding absolute file paths: data in same directory

- If you data is in the same directory as your code (or a subdirectory), the `here` package is your friend
- `here::here("data", "inputs.csv")` will find the "inputs.csv" file in the data subdirectory of your project, regardless of where the R script is saved

## Avoiding absolute paths: data in different directory

- If your data is in a different directory, you can use a `.Renviron` file to specify a file path
- Let's create a `.Renviron` file in our package
- Add to the file

```
DATA_PATH="path/to/OdumModularDesignR/data"
```

## Accessing the directory

- Within R, you can get the data path from `.Renviron` with `Sys.getenv("DATA_PATH")`
- At the top of my script I usually have something like
    ```r
    DATA_PATH = Sys.getenv("DATA_PATH")
    ```
- Whenever I read a file, I have
    ```r
    data = read_csv(file.path(DATA_PATH, "file_name.csv"))
    ```

## Using .Renviron in a team

- This is particularly useful when working with a team
- _Don't_ check `.Renviron` into version control
    - Better yet, add it to `.gitignore`
- Each team member creates their own `.Renviron` with where the data is _on their computer_
- All the other scripts work on everyone's computer without modification
- I often create a `.Renviron.template` that I check in to version control to show folks what variable(s) need to be set
    - Setting up on a new computer: copy `.Renviron.template` to `.Renviron` and edit the included paths

## Using functions in multiple files

- Often we may want to use functions in multiple files
- The best way to do this is to create a _package_
- You've used lots of packages in R - anytime you use `library()` you're loading a package

## Creating R packages

- You use packages all the time when you use R
    - Any time you run `library(...)` you're loading a package
- Creating your own package is very simple, once you know how to do it
- I structure almost every analysis I do as a package - anything more complicated than a single script
    - [A recent paper](https://github.com/mattwigway/BetweennessCentralityResilience/)
    - [A package designed for reading transit data](github.com/mattwigway/gtfsrealtime-r)
- R packages are great for sharing code with others, but I use them even on solo projects

## Advantages of creating a package

- Allow sharing code between analysis files
- Facilitate automated testing
- Facilitate versioning

## Creating an R package

- Several packages are available that make creating your own packages _much_ easier
    - `devtools` for overall workflow management
    - `roxygen2` for documentation generation and 
    - `testthat` for automated testing
- Install these now
```{r}
#| eval: false
install.packages(c("devtools", "roxygen2", "testthat"))
```

## What is an R package?

- A collection of functions, data, and documentation that can be easily installed in R and available to all R scripts

## Structure of an R package

- R packages are created in _source_ form
- This just means that they consist of the original R code
- When you install a package, it is converted to _binary_ form
    - Only readable by the computer, and ready to be run by R

## Structure of an R package

An R source package is just a directory with a specific structure

- `packagename`
    - `packagename.Rproj`
    - `.Rbuildignore`
    - `DESCRIPTION`
    - `NAMESPACE`
    - `R`
        - Your R files

## Creating your R package

- Use `usethis::create_package(...)` to create the directory structure for your package
- `usethis::create_package(...)` will create the directory if it doesn't already exist
- Since we're likely only going to use this package in this project, we can create the package in the same directory
    - if we were sharing it across many projects, we'd make the package itself a separate directory/project
- `usethis::create_package(".")` to create a package in the current directory

## Adding R code to your package

- Your R code goes in files in the `R/` directory
- Any file in this directory ending in `.R` will be part of your package

## Sidebar: namespaces

- Multiple packages may define functions with the same name
- To keep this straight, R uses _namespaces_
- Each package gets a namespace, and functions are defined within that namespace
- When you use `library(package)`, functions from that package are loaded into the global (main) namespace
- `package::function` refers to a function within a specific package's namespace

## Differences between R scripts and package code

- Packages should never use `library()` as this could affect code from other packages
- Almost everything in a package should be inside a function
    - Code that is not in functions will be run when the package is compiled or installed, not when it is used
    - Main exception is constants

## Creating a data cleaning function

- Create a function that takes arguments for the file names of the dataset and the two tables that are joined to it,
- performs the data cleaning/processing,
- and returns the final dataset
    - If the final dataset is not the result of the last line of your function, add `return(data)` at the end to return `data` (or whatever your dataset is called)
- Save this function to a new file in the same directory as the others


## The finished function

Yours might not look exactly the same, and that's okay

```{r}
#| output: false
load_air_data = function (filename, cities, carriers) {
  # first, we need to load the data
  data = read_csv(filename)
  
  # The data have seven columns: origin and destination airport, origin and destination cities
  # carrier, and distance. The city and carrier are coded, so we will merge in other
  # (the airports have codes as well, but these are fairly well known - e.g. RDU is
  # Raleigh-Durham and LAX is Los Angeles; we won't match those with the official airport
  # names)
  
  market_ids = read_csv(cities)
  data = left_join(data, rename(market_ids, OriginCity = "Description"), by = c(OriginCityMarketID = "Code"))
  data = left_join(data, rename(market_ids, DestCity = "Description"), by = c(DestCityMarketID = "Code"))
  
  carriers = read_csv(carriers)
  data = left_join(data, rename(carriers, OperatingCarrierName = "Description"), by = c(OpCarrier = "Code"))
  data = left_join(data, rename(carriers, TicketingCarrierName = "Description"), by = c(TkCarrier = "Code"))
  
  return(data)
}
```

## Using your data cleaning function

- Delete the data loading code in each of the analysis files
- Use the 🧹 button in RStudio to clear the data from memory
- Add and run a line `devtools::load_all()` to load your new package
- Call your new function like this:

```{r}
#| eval: false
DATA_PATH = Sys.getenv("DATA_PATH")
data = load_air_data(file.path(DATA_PATH, "air_sample.csv"), file.path(DATA_PATH, "L_CITY_MARKET_ID.csv"), file.path(DATA_PATH, "L_CARRIERS.csv"))
```

## Oops, error

- You probably got this: `Error in read_csv(filename) : could not find function "read_csv"`
- Normally, `read_csv` and other functions become available when you run `library(tidyverse)`
- But we don't do this in libraries
- There are two main ways to refer to functions from other libraries
    - Fully-qualified notation: `readr::read_csv`
    - `@importFrom` comments


## Fully-qualified notation

- When using fully qualified notation, we put the name of the library followed by `::` in front of the function name
- So, `read_csv` is replaced with `readr::read_csv`
- `left_join` and `rename` are prefixed with `dplyr::`

## Fully qualified notation

- We tell R that we want to use functions from the libraries `readr` and `dplyr` by running `usethis::use_package()`
- This ensures that these packages are listed as _dependencies_, so when we tell R to install the package it ensures these are installed as well.
- Do this, and run `devtools::load_all` again, and confirm that `load_air_data` now works

```{r}
#| eval: false
usethis::use_package("readr")
usethis::use_package("dplyr")
```

## Confirm it worked

- Re-run `load_all`
- Run the data loading

```{r}
#| eval: false
data = load_air_data(file.path(DATA_PATH, "air_sample.csv"), file.path(DATA_PATH, "L_CITY_MARKET_ID.csv"), file.path(DATA_PATH, "L_CARRIERS.csv"))
```


## Using `tidyverse` functionality

- `tidyverse` is a collection of many packages
- Using `tidyverse` directly in libraries [is discouraged](https://tidyverse.org/blog/2018/06/tidyverse-not-for-packages/)
- Most data manipulation functions are from `dplyr`, data reading functions from `readr`
- Loading `tidyverse` in the console and running `?function` will tell you what constituent package it comes from

## `@importFrom`

- If you don't want to type out `dplyr::` over and over, you can use `@importFrom` to bring a function into your package's namespace
- The `usethis::use_import_from` function automates this task

## `@importFrom`

- Let's import `summarize`, `group_by`, `mutate`, `first`, `if_else`, `arrange`, `ungroup`, `filter` and `summarize` from `dplyr` to make our analysis code work

```{r}
#| eval: false
usethis::use_import_from("dplyr", c("summarize", "group_by", "mutate", "first", "if_else", "filter", "ungroup", "arrange", "summarize"))
```

- Accept the prompt to generate package documentation

## What happened?

- `devtools` created a new `airlinestats-package.R` that contains `@importFrom` statements to bring these functions into the package namespace
- The file consists of comments, but the `roxygen2` documentation system will convert these comments into import statements in the file `NAMESPACE` when the documentation is built

## Adding R code to your package

- Create three files, `busiest_routes.R`, `market_shares.R`,  and `constants.R`
- Put the functions to calculate market shares and route popularity in `market_shares.R` and `busiest_routes.R`, respectively
    - You can have multiple functions per file, and the names don't have to match the filename, but I try to keep distinct analyses
        organized into different files
- Put the symbolic constant `KILOMETERS_PER_MILE` in `constants.R`
    - Good practice to put your constants in one file so they don't get defined in multiple places accidentally
- _Don't_ include `library(tidyverse)`
- Save the files in the `R` directory of your package 

```{r}
#| eval: false
market_shares(data, OperatingCarrierName, OriginCity)
busiest_routes(data, OriginCity, DestCity)
```

## Checking the package

- Run the `devtools::check()` function

## Checking the package

```{r}
#| eval: false
check()
```

. . .

<pre>
❯ checking DESCRIPTION meta-information ... WARNING
  Non-standard license specification:
    `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license
  Standardizable: FALSE

❯ checking R code for possible problems ... NOTE
  busiest_routes: no visible binding for global variable ‘Passengers’
  busiest_routes: no visible binding for global variable ‘Distance’
  busiest_routes: no visible binding for global variable ‘airport1’
  busiest_routes: no visible binding for global variable ‘airport2’
  busiest_routes: no visible binding for global variable ‘distance_km’
  market_shares: no visible binding for global variable ‘Passengers’
  market_shares: no visible binding for global variable ‘market_share’
  Undefined global functions or variables:
    Distance Passengers airport1 airport2 distance_km market_share
N  checking for hidden files and directories ...
   Found the following hidden files and directories:
     .quarto
   These were most likely included in error. See section ‘Package
   structure’ in the ‘Writing R Extensions’ manual.
N  checking for portable file names ...
   Found the following non-portable file paths:
     OdumDemo/docs/index_files/libs/quarto-html/quarto-syntax-highlighting-46f66b053e56abd73bb6e954d61a315b.css
</pre>

## Addressing the issues

- `Non-standard license specification`
    - R is suggesting you fill out the `DESCRIPTION` file, which includes choosing a license. We'll skip this for now.
- `No visible binding` and `undefined global`
    - These are the columns of our data; R doesn't know this when checking the code, though, and warns that they may be undefined
    - This is fine for packages used internally; for packages you are planning to publish you often wouldn't want to assume certain columns were in the data
- Non-portable file paths, non-standard files, and hidden directories - these are most often files that you use in development that shouldn't be part of the package anyways
    - `air-analysis.R` uses the package but shouldn't be part of it

## Ignoring unwanted files

- Like `.gitignore`, there is an `.Rbuildignore` to ignore files when building an R package
- Easiest way is again through `usethis`:

```r
usethis::use_build_ignore("docs/")
usethis::use_build_ignore(".quarto")
usethis::use_build_ignore("_quarto.yml")
usethis::use_build_ignore("slides")
usethis::use_build_ignore("data/")
usethis::use_build_ignore("answers/")
usethis::use_build_ignore("index.qmd")
usethis::use_build_ignore("air_analysis.R")
usethis::use_build_ignore("airline_analysis.R")
```

- Run `devtools::check()` again

## Bringing it full circle

- Return to the original script files
- Replace the code in `air_analysis.R` and confirm you get the same results

```{r}
#| eval: false
library(tidyverse)
devtools::load_all()

DATA_PATH = Sys.getenv("DATA_PATH")

# first, we need to load the data
data = load_air_data(file.path(DATA_PATH, "air_sample.csv"), file.path(DATA_PATH, "L_CITY_MARKET_ID.csv"), file.path(DATA_PATH, "L_CARRIERS.csv"))

busiest_routes(data, Origin, Dest)

# This may be misleading, however, as some metropolitan areas have only one airport
# (for example, Raleigh-Durham or Las Vegas), while others have more (for example,
# New York or Los Angeles). We can repeat the analysis grouping by "market", which
# groups these airports together.
# Now, we can see what the most popular air route is, by summing up the number of
# passengers carried.
busiest_routes(data, OriginCity, DestCity)
```

## Automated testing

- Automated testing formalizes testing and makes it repeatable
- Every test is run automatically when you run `check()`
- If you break something later, you are more likely to find out about it if your code has automated tests

## The structure of testing

- Automated tests consist of code and _expectations_
    - Expectations are called _assertions_ in other languages
- A test exercises one of the functions in your package, and specifies expectations about its output
- Good tests are small, with one piece of code tested and a handful of expections
- This may require splitting your package code up into multiple functions
- It's a good idea to test how your package acts when used incorrectly as well—for instance, when fed bad data

## Automated testing in R

- In R, automated testing is provided by the `testthat` package
- We can add testing to our package by running

```{r}
#| eval: false
usethis::use_testthat(3)
```

## Automated testing in R

- We now have a `tests/` directory, with a `testthat` directory inside it
- Any file starting with `test` and ending with `.R` in this directory will be run as part of automated testing

## Writing tests

- Create an R file to hold tests for the analysis functions
- It's good practice to match the names of the test files to the names of the files that define the code being tested, so create `tests/testthat/test-market-shares.R`

## Writing tests

- We'll write two tests for the analysis code
    1. We'll test that market shares of all airline at an airport sum to 1
    2. We'll test that passing in a dataset without a `Passengers` column leads to an error

## Writing tests

- Often, we need data for our tests
- Here, we will create a small dataset within our test file directly
- It is also possible to distribute a CSV with your package with test data
- I like to put my code to generate test data in a function, and re-create the data for each test, so that there is no chance of one test modifying the data used by another
    - Unless the data take so long to create or load that this is impractical

## Test data

```{r}
#| eval: false
get_data = function () {
  data = dplyr::tibble(
    Origin=rep(c("SFO", "ORD", "DCA", "PHX", "BOS", "RDU"), 20),
    Destination=rep(c("DFW", "MCI", "MIA", "ABQ", "IAH", "SJU"), 20),
    Passengers=rep(1:6, 20),
    Carrier=rep(c("United", "American", "Delta", "Southwest"), 30)
  )
  return(data)
}
```

## Write the test for market shares

```{r}
#| eval: false
test_that("Market shares sum to 1", {
  data = get_data()
  shares = market_shares(data, Carrier, Origin)
  total_shares = group_by(shares, Origin) |> summarize(overall=sum(market_share))
  # six airports, so six shares
  expect_equal(total_shares$overall, rep(1, 6))
})
```

## Write the test for bad input

```{r}
#| eval: false
test_that("Error when passengers column missing", {
  data = get_data() |> dplyr::select(-Passengers)
  expect_error(busiest_routes(data, Origin, Destination))
})
```

## Run tests

- Click "run tests" in the upper-right corner in RStudio
    - This will run the tests in the current test file
- Alternately, run
```{r}
#| eval: false
devtools::test()
```

to test everything

## Fix test errors

- We got a test failure for the market shares test
- This is because we filtered out small airlines in the function
- Let's remove the filtering from the function
    - It's good to break your functions into smaller pieces for easier testing

## Exercise

- Check and reinstall your package
- Notice that your tests were run
- Confirm your scripts still work

## Including data in tests

- Sometimes, you'll need to include toy datasets to use with your tests
- Let's write a new test for the `busiest_routes` function, to make sure it identifies the busiest routes
- We'll make a toy dataset to work with

## Including data in tests

- Put data files in `inst/testdata`
- Call it `passengers.csv`

```
from,to,Passengers,Distance
SFO,JFK,2,2586
RDU,CLT,3,130
JFK,SFO,2,2586
```

## Writing the test

- Write a test that reads the data in and tests that `busiest_routes` returns JFK->SFO as the busiest route
    - Note that I have cleverly also tested that both directions are combined, because SFO->JFK and JFK->SFO are individually smaller than RDU->CLT
- You can get the path to your `passengers.csv` file with `system.file("testdata", "passengers.csv", package="airlinestats")`

## Result

```{r}
#| eval: false
test_that("busiest_routes works", {
  data = readr::read_csv(system.file("testdata/passengers.csv", package="OdumModularDesignR"))
  result = busiest_routes(data, from, to)
  expect_equal(result$airport1, c("JFK", "CLT"))
  expect_equal(result$airport2, c("SFO", "RDU"))
  expect_equal(result$Passengers, c(4, 3))
})
```

## Environment management and `renv`

- R projects rely heavily on outside packages
- Changes to package versions may break code or change results
- Reproducibility requires keeping track of used packages and versions

## `renv`

- `renv` is a package that automates this
- Once installed, set up a new project with `renv::init()` and snapshots current packages in use in a _lockfile_
- `renv::snapshot()` updates currently used packages
- `renv::restore()` will reinstall dependencies on a new machine or years in the future

## Exercise: set up `renv` in your project

- Install `renv` if needed
- Run `renv::init()`
- We will run `renv` in "implicit mode" - this way it will track packages used in our scripts
- "Explicit" mode only tracks the packages you've added with `use_package` or `use_import_from`
- Inspect `renv.lock`

## `renv.lock`

- Records the exact versions of all packages used
- Allows `renv` to reproduce the environment exactly on another machine or in the future

## `renv.lock`

{{< include renvlock.qmd >}}

##  `renv` environments

- Your `renv` environment is specific to each project
- Installing or upgrading a package in a project managed by `renv` will not affect any other projects on your computer

## Updating an `renv` environment

- You can install packages in `renv` just like you normally would, with `install.packages`
    - `renv::install` works as well
- When you add or remove packages, run `renv::snapshot` to update the lockfile
- By default, the lockfile will reflect all the packages you use in your code (via `library()`, `package::method()`, etc.), and all the packages they depend on
    - Won't include installed but unused packages

## Restoring an environment

- `renv::restore()` will read a lockfile and (try to) restore exactly the packages used
    - This may fail if the packages are no longer available
    - CRAN does not keep older binary packages, so older packages may need to be built from source
- Don't do this now, since your environment is already set up

## `renv` and package development

- If you're keeping your scripts in your package repository, feel free to use `renv`
- If your package is used by multiple projects, you want to use `renv` with each individual project
    - Those projects might require additional dependencies
    - Packages intended for reuse will need to coexist with other packages - if two packages used renv and requested different versions of (say) dplyr, that would be an issue
- This is why explicit mode is useless IMHO - the only reason to do it would be if your package was intended to be distributed to others, and
    in that case `renv` should be set up in their projects, not in the package

## `renv` and a project-based workflow

- `renv` encourages you to think of your R projects as separate, logical entities
- If you're using `renv`, you can't ever switch projects without also restarting R
    - This is a good thing
- Using a [project-based workflow](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) lets you make your code portable between machines
- Key points: avoid `setwd`, `rm(list=ls())`, and saving `.RData`
    - You can even configure RStudio to never save `.RData`, which I recommend

## `renv` and version control

- You should check `renv.lock` and your `renv/` directory into version control
    - not `renv/library` if you have it, this contains the actual packages
    - should be ignored automatically by Git

## Getting our package on Github

- Since we originally cloned the repository from Github, if we try to push it it will just try to push to the original repository
- To "disconnect" the repository from the original Github repository:
    - `git remote rm origin`
- Then, create a new Github repo
- Commit the changes you've made
- And push the code to it

## Continuous integration

- We've been writing automated tests, but we have to run them manually
- Realistically, you're probably not going to do this very often
- You could miss broken tests for a long time
- If you're using Github, we can set it up to run your tests on every push

## Continuous integration

- This is particularly valuable if you're using branches for development (you should)
- When you create a pull request, [Github will display whether the tests pass before you merge](https://github.com/mattwigway/gtfsrealtime-r/pull/14)

## Setting up continuous integration

- Continuous integration is set up through [Github Actions](https://github.com/actions)
- You add a workflow file in the `.github/workflows` directory

## Setting up continuous integration

- `usethis` provides continuous integration templates
- Run `usethis::use_github_action("check-release")`
- Inspect the file in `.github/workflows`

## Setting up continuous integration

- Commit your `.github` directory and push to github

## See if it worked

- Browse to your repository on Github. Look for the brown dot to indicate workflows are running to the right of the commit message.

## Two more warnings

- We still get a few warnings
- We want to ignore those for now, though this is not good practice in general
    - Better to fix them
- In your `check-release.yaml` file, add this below the line that says to use `check-r-package`:
    - ` error-on: '"error"'
- The quote order is important: double inside single

## Using `renv` with CI

- To have Github install the exact packages we're using into the CI environment, we need to make one change
- Swap out the `setup-r-dependencies` section of your workflow for `setup-renv`
- No need to include `with: ` or the variables below it
- Commit and push

## CI and branch-based development

- CI is especially useful working with a team
- CI will run on each pull request, and show if tests pass

## Documentation

- R previously complained that we didn't have documentation for our exported functions
- It's a good practice to document all of our externally-facing functions (and required, if you plan to submit to CRAN)
- The roxygen2 system allows us to do this right in our R code

## Roxygen2 comments

```{r}
#' Compute carrier market shares by airport
#'
#' This function computes the market share of each carrier at each airport, with options for determining
#' how to determine the carrier and the airport (e.g. ticketed vs. operating, individual airport vs. metro
#' area airport system).
#'
#' The carrier and airport columns can be specified manually, so ticketing or operating
#' carrier can be used (or another carrier column, if you had one---for instance, mainline
#' vs regional carrier market share, or legacy vs. low-cost). The airport column is likewise
#' an argument, so market shares per-airport, per-city, per-state, etc. can be calculated.
#'
#' @param dataframe data to use
#' @param carriercol column of data frame that specifies carrier
#' @param origincol column of data frame that specifies airport
#' 
#' @examplesIf require("tibble")
#' data = tibble::tribble(
#'     ~Origin, ~Origin_City, ~Dest, ~Passengers, ~Carrier,
#'     "LAX",   "Los Angeles", "SFO", 400,        "DL",
#'     "LAX",   "Los Angeles", "IAD", 200,        "UA",
#'     "LGB",   "Los Angeles", "SJC", 100,        "B6"
#' )
#' 
#' # compute market shares for departures from each airport
#' market_shares(data, Carrier, Origin)
#' 
#' # compute market shares for departures combing all airports in Los Angeles area
#' market_shares(data, Carrier, Origin_City)
#'
#' @returns dataset containing market shares, with one row per by airline and airport
#'
#' @export
```

## Roxygen2 comment anatomy: title

```{r}
#' Compute carrier market shares by airport
```

## Roxygen2 comment anatomy: description

```{r}
#' This function computes the market share of each carrier at each airport.
```

## Roxygen2 comment anatomy: details

```{r}
#' The carrier and airport columns can be specified manually, so ticketing or operating
#' carrier can be used (or another carrier column, if you had one---for instance, mainline
#' vs regional carrier market share, or legacy vs. low-cost). The airport column is likewise
#' an argument, so market shares per-airport, per-city, per-state, etc. can be calculated.
```

## Roxygen2 comment anatomy: arguments (parameters)

```{r}
#' @param dataframe data to use
#' @param carriercol column of data frame that specifies carrier
#' @param origincol column of data frame that specifies airport
```

## Roxygen2 comment anatomy: return value

```{r}
#' @returns dataset containing market shares by airline and airport
```

## Roxygen2 comment anatomy: examples

- Everything between `@examples` and the next directive is R code for examples
- This code will be run when building your package, so it should work...
- If it requires packages, etc., `@examplesIf` will run the code only if whatever is immediately after `@requiresIf` is true
- Examples must be self-contained - i.e. run without any other code (other than loading your library) needing to be run first

## Roxygen2 comment anatomy: examples: example


```{r}
#' @examplesIf require("tibble")
#' data = tibble::tribble(
#'     ~Origin, ~Origin_City, ~Dest, ~Passengers, ~Carrier,
#'     "LAX",   "Los Angeles", "SFO", 400,        "DL",
#'     "LAX",   "Los Angeles", "IAD", 200,        "UA",
#'     "LGB",   "Los Angeles", "SJC", 100,        "B6"
#' )
#' 
#' # compute market shares for departures from each airport
#' market_shares(data, Carrier, Origin)
#' 
#' # compute market shares for departures combing all airports in Los Angeles area
#' market_shares(data, Carrier, Origin_City)
```

## Roxygen2 comment anatomy: export

```{r}
#' @export
```

- This indicates which functions will be loaded when you run `library(package)`
- It is optional when using devtools, but good practice to clarify which functions are intended for use outside the package
- It is required if you are distributing your package to be used in other projects

## Examples and `Suggests`

- In our example above, we used the `tibble` package
- This package is not necessary to use our package, but it is necessary to run the examples
- R has a special way to depend on these - the `Suggests` field in the `DESCRIPTION` file
- Let's add the tibble package to that field

## Other Roxygen2 directives

- `@noRd` - do not show this in the user facing documentation (useful for documenting internal functions)

## Building the documentation

- Add a roxygen2 comment to the `market_shares` function
- Build documentation with `document()`
- Notice that a new folder, `man` (manual), has appeared
- Reload package with `devtools::load_all()`
- Run `?market_shares`

## Exercise

- Document `busiest_routes` and `load_air_data`
- Build the documentation and reload the module

## The `man` folder and git

- It's typical to commit the `man` folder to git, which means you need to run `document()` before committing changes to your repository

## Sharing your package with the world

- The easiest way to distribute your package is to just have it on Github
- Others can install it with `renv::install` or `devtools::install_github`
    - Then it can be loaded with `library(package_name)`
    - Remember to mark functions with `@export`
- For packages that you expect to be of broad interest, CRAN is a good option
    - The repository that `install.packages` installs packages from
    - Notoriously unpleasant and nitpicky
    - Read [the book](https://r-pkgs.org/release.html) before you do this
- For packages with compiled C/C++/Rust code, [R-universe](https://r-universe.dev/search) is a good option
    - If you don't know what this means, you don't have compiled C/C++/Rust code

## Semantic versioning

- Modern open-source software across many languages is switching to a common versioning scheme known as [_semantic versioning_](https://semver.org/)
- Version numbers are three numbers separated by periods: the major version, the minor version, and the patch version
- The patch version is incremented when you release a new version that only contains bugfixes
- The minor version is incremented when you release a new version with new functionality that is backwards-compatible (i.e. someone using version 1.2 can update to version 1.3 without needing to change their code)
- The major version is incremented when you release a new version with "breaking changes" - i.e. upgrading from 1.0 to 2.0 might mean other projects need to update their code
- For version 0.x.y, anything goes - but typically folks will increment the minor version for breaking changes and the patch for anything else

## Release versions

- When you release a version, it's a good idea to:
    - Update the version number in `DESCRIPTION`
    - Create a tag on Github (typically the tag would start with a `v`, e.g. `v1.0.0`)
    - Create a release on Github from that tag
    - Have a changelog describing the change (typically in NEWS.md).
    - When under development, it is typical for the version to be set to whatever the previous version was, with `.9000` added to the end
- Then people can install a specific version with `devtools::install_github("user/package@v1.0.0")` or `renv::install("user/package@v1.0.0")`

## Building a website for your package

- The best way to distribute documentation is through a [pkgdown](https://pkgdown.r-lib.org/) site
    - For example, [this](https://projects.indicatrix.org/gtfsrealtime-r/) is the pkgdown site for one of the packages I maintain
- The easiest way to set this up is `usethis::use_pkgdown_github_pages()`
- This will make your pkgdown site available at `https://<username>.github.io/<package>`
    - **This will be public even if your repository is private**

## Vignettes

- The other common form of documentation for R packages is a vignette
- Prose documents written as R markdown files

## Markdown

- Markdown is a simple text formatting language
```
## Heading 1
### Heading 2
####### Heading 6

_italic_
**bold**

[link text](link url)
```

## R Markdown

- R Markdown allows embedding R code in markdown, between ` ```{r} ` and ` ``` `
- Embedded R code will be executed, and outputs placed in the document

## Writing a vignette for our package

- Run `usethis::use_vignette("airlinestats")` to create the vignette
    - "airlinestats" is the name of the vignette. You might have many vignettes with different names, rather than a single one matching the name of your package

## Writing a vignette for our package

```
---
title: "market_share_calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{market_share_calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```

## Writing text

- You can write text simply as text - with markdown formatting as needed

## Including code in the vignette

- Include the following code in your vignette, surrounded by ` ```{r} ` and ` ``` `

```{r}
data = tibble::tibble(
    Origin=c("RDU", "RDU", "CLT", "CLT", "CLT"),
    OperatingCarrierName = c("Southwest", "Delta", "American", "American", "Piedmont"),
    Passengers=rep(1, 5)
)

market_shares(data, Origin, OperatingCarrierName)
```

## Previewing your vignette

- I find the easiest way is to just swap out `library()` for `devtools::load_all()` and then use the RStudio render button

## Using your vignette

- Once your package is installed, you can also see the vignette with `vignette("vignette_name", package="package_name")` or `browseVignettes("packagename")`
- Vignettes will also show up on your `pkgdown` site

## Fix the error

- Add `- uses: r-lib/actions/setup-pandoc@v2` to your workflow to allow it to build your RMarkdown vignettes

## Vignettes and installing from Github

- Vignettes are not built when installing from github, to save time
    - Most packages will have their vignettes available on a `pkgdown` site anyhow
- Vignettes are mostly for packages you expect many others to use

## Fixing the license

- The last warning we had from the CI was that our package uses a non-standard license
- There are two ways to specify a license in R
- A [standard license](https://usethis.r-lib.org/reference/licenses.html), which is a code for one of several common Open Source licenses
- A custom license, which might be "all rights reserved" or some other open-source license not specified

## Fixing the license

- We'll use a proprietary license for now
    - Or if you prefer an open source one you can do that, e.g. `usethis::use_mit_license()`
- `usethis::use_proprietary_license("your name")`

## Additional resources

- R package book: [r-pkgs.org](https://r-pkgs.org/)
- R Github Actions documentation: [github.com/r-lib/actions](https://github.com/r-lib/actions)

### Contact

Matt Bhagat-Conway<br/>
[mwbc@unc.edu](mailto:mwbc@unc.edu)